---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mevolCVP2

<!-- badges: start -->
[![R-CMD-check](https://github.com/vbonhomme/mevolCVP2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/vbonhomme/mevolCVP2/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of mevolCVP2 is to provide both a general approach and helpers to parameterize Linear Discriminant Analyses. It generalizes the seminal idea [first published](https://www.sciencedirect.com/science/article/abs/pii/S030544031200355X) in 2013 by Allowen Evin and colleagues in _Journal of Archaeological Science_, and follows a pre-package version by [Evin and Bonhomme](https://github.com/vbonhomme/mevolCVP).

This approach was initially used for morphometrical analyses but may reveal useful in other fields. For morphometrics, a massive user of Linear Discriminant Analyses.

Yet this approach deeply changed the way we see, parameterize and do LDAs in morphometrics, it boils down, programmatically, to very few helpers that we propose here.

The vignette presents more extensively the package architecture as well as ways to flavour it upon your own needs.

## Installation

You can install the development version of mevolCVP2 from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("vbonhomme/mevolCVP2")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(mevolCVP2)
## basic example code
```
